### 2009-05-27 paper.tex
more edits

Consider the stochastic grammar which generates parse trees from the marginal distribution $P(X)$.
It is occasionally convenient to represent this grammar as a transducer whose input is constrained to be null,
i.e. a machine that accepts a dummy (empty) input sequence, and outputs sequence $X$.
We refer to this as the {\bf singlet transducer}.
In contrast, the more general type of transducer that absorbs parse trees $X$ and generates modified parse trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
By definition, singlet transducers only emit symbols to their output sequence, and use a restricted set of state types.
Branch transducers, in contrast, can both emit symbols to their outputs and absorb symbols from their inputs, and so use the full range of state types.

@@

The model construction algorithm described above creates a phylogenetic SCFG
which generates a joint distribution over both ancestral and extant
sequences. A parse tree for this SCFG represents a structural and evolutionary
explanation of the extant sequences, including a complete ancestral reconstruction.
Therefore, given a set of extant sequences, many of the questions of interest to us
can be reduced to searches over, or summarizations of, the set of possible parse trees.

Well-known algorithms exist for searching or summing over SCFG parse trees.
The Cocke-Younger-Kasami (CYK) algorithm performs maximum-likelihood (ML) inference;
the Inside algorithm can be used to sum over parse trees or sample them {\em a posteriori};
and the Inside-Outside algorithm yields posterior probabilities for individual parse tree nodes.

All of these algorithms are, however, complicated for phylogenetic SCFGs by
the possibility of fragments of ancestral sequence that were universally deleted and therefore
are unobserved in any of the extant sequences.
These unobserved fragments can be unbounded in length.
In the parse tree, these fragments correspond to cyclic detours through $\Snull$ states,
along with bifurcations where one or both sides terminate without observable emissions
(e.g. $B \to X\ Y$ followed by $X \to \epsilon$).
In the CYK, Inside and Outside recursions, this causes cyclic dependencies which cannot be resolved.

Therefore, before presenting equivalents of the CYK, Inside and Outside algorithms,
we first give a method to eliminate ``null cycles'' from
the ensemble model, by transforming any SCFG to an equivalent acyclic SCFG.
Some such step is required, in order to deal with cyclic dependencies, but alternative approaches to ours are clearly possible.
A CYK-only approach appears in the computational linguistics literature \cite{GecsegSteinby97}.
We have also developed an independent heuristic that iteratively re-fills entries of the DP matrix
until the approximation is sufficiently good.

\subsubsection*{Eliminating null cycles in SCFGs}

@@

In contrast, an {\bf empty path} is defined as any parse-subtree {\em without} bifurcations that does not emit terminal symbols.
Most common are paths of ``windback'' null transitions via non-emitting placeholder states.
If $\Semit$ states $E_1$ and $E_2$ are connected in the state
graph via $\Snull$ states $N_1$ and $N_2$, then the path $E_1
\rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
$P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
\rightarrow E_2)$ can be replaced by a single direct transition $E_1 \rightarrow
E_2$ with an identical probability.

@@

Empty paths and empty bifurcations can be combined to give finite-probability sequences of rules
that transform a nonterminal back into itself, with no observable emissions.
We refer to such sequences of rules as {\bf null cycles}.
Our goal is an algorithmic procedure to account for the likelihood of such cycles by exact marginalization.
For simpler models, solutions to this problem are published.
Eliminating missing (empty) columns from an independent-sites model of sequence involves a straightforward application of a correction factor
to account for the proportion of columns that are unobserved \cite{RivasEddy2008}.
The slightly more complicated situation of missing emissions in a HMM can be dealt with by
summing over all empty paths, yielding a geometric series that is solvable by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007}.

Algorithms for exact elimination of empty paths from HMMs by matrix inversion are implemented
by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and
HMMoC \cite{Lunter2007}. These algorithms effectively replace the HMM
with an equivalent HMM that contains no null cycles but generates the
same probability distribution over sequences.
The ``wing retraction'' algorithms used in the repetitively structured profile HMMs of HMMER \cite{hmmer} appear similar in operation.
However, these solutions do not easily generalize to SCFGs, which contain bifurcations
(and, therefore, may generate empty bifurcations as well as empty paths).

@@

We give a formal description of the null-cycle elimination procedure,
along with procedures for probabilistically ``restoring'' null cycles
to sampled parse trees and Inside-Outside expectation counts, below.

@@

Exact inference on a star phylogeny with $N$ extant sequences therefore
has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in space and
time for a multiple-SCFG with $A$ states and $B$ bifurcations. As
described earlier, in practice we frequently have expert knowledge
(such as a curated multiple alignment)
about the structures and/or evolutionary histories of the sequences of
interest.  We can use this knowledge as a constraint to reduce the accessible volume of the DP matrix \cite{Holmes2005}.
Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
Outside and CYK traceback algorithms for a three-taxon star phylogeny,
constrained using the ``fold envelope'' concept.

@@

We illustrated our method for building models of structured sequences
using a simplified model which was introduced in previous work,
the TKF Structure Tree model \cite{Holmes2004}, a simple probabilistic
model of the evolution of RNA structure.

@@

The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).

@@

% Prior art
The work reported in this paper builds on substantial prior art in the areas of evolutionary modeling and ancestral reconstruction.
The reconstruction of ancient sequences was first proposed in 1963 by Pauling and Zuckerkandl \cite{PaulingZuckerkandl63};
current applications of this idea, mostly using substitution models, are surveyed in the book edited by Liberles \cite{Liberles2007}.
Many algorithms in phylogenetics implicitly reconstruct substitution histories,
whether by parsimony \cite{Edwards63,HendyPenny82} or likelihood \cite{Felsenstein81}.
There has been much effort to model the evolution of sequences related
by a tree with models permitting indels
\cite{ThorneEtal91,HeinEtal2000,Hein2001,HolmesBruno2001,SteelHein2001,Holmes2003,KnudsenMiyamoto2003,LunterSongMiklosHein2003,MiklosLunterHolmes2004,LunterEtAl2004,BradleyHolmes2007,SatijaEtAl2008}.
Recent work has extended these ideas to the reconstruction of indel
histories \cite{KimSinha2007,DialloEtAl2007},
particularly at the genomic scale \cite{MillerHausslerEtAl2006,PatenHolmesBirney2008}.
There is substantial prior work in computational linguistics
on the theory of transducers for sequences \cite{MohriPereiraRiley2000}
and parse trees \cite{Rounds70,Thatcher70,GecsegSteinby97,ComonEtAl2007-TreeTransducers}
(from which we take the terms ``string transducer'' and ``parse-tree transducer'').
We draw on the bioinformatics literature for SCFGs
\cite{Eddy94,Sakakibara94c,Durbin98},
especially Pair SCFGs \cite{RivasEddy99,Holmes2005,DowellEddy2006}
and phylogenetic SCFGs \cite{KnudsenHein2003}.
In particular, an early example of a pairwise conditional model $P(Y|X)$ for structure-dependent RNA evolution
may be found in \cite{KleinEddy2003}.
A conditional framework similar to ours in some respects is described
by Sakakibara et al \cite{Sakakibara2003}.
The dynamic programming inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85}.
%a seminal work in both RNA structural alignment and ancestral reconstruction.

@@

While we have focused on the TKF Structure Tree model in our Results,
our model-construction algorithm is applicable to any model of the evolution of secondary
structure which can be expressed as a Pair SCFG.
Realistic structural and thermodynamic effects---such as base-stacking
or loop length distributions---can, in principle, be incorporated.
%Furthermore, while an evolutionary model of structural evolution
%is necessarily complex,
%our model-construction algorithm enforces a compact parametrization,
%allowing fast training on structural alignments of pairs of sequences.

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of those currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
Thus, alternate strategies will have to be explored, such as the
progressive reconstruction used by the \ortheus\ program for
reconstructing ancestral genomes \cite{PatenHolmesBirney2008}.
Another promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

@@

In summary, the evolutionary models we have described form a theoretical
foundation for studying the structural evolution of RNA gene families
in detail. By opening the door to ancestral sequence reconstruction
and paleogenomics, these algorithms will allow us to test our
knowledge of RNA evolution by direct experimental investigation
of early ribonucleic machines.

>>>>

Consider the stochastic grammar which generates parse trees from the marginal distribution $P(X)$.
It is convenient to represent this grammar as a transducer whose input is constrained to be null,
i.e. a machine that accepts a dummy (empty) input sequence, and outputs sequence $X$.
We refer to this as the {\bf singlet transducer}.
In contrast, the more general type of transducer that absorbs parse trees $X$ and generates modified parse trees $Y$ from the conditional distribution $P(Y | X, T)$ is a {\bf branch transducer}.
By definition, singlet transducers only emit symbols to their output sequence, and use a restricted set of state types.
Branch transducers, in contrast, can both emit symbols to their outputs and absorb symbols from their inputs, and so use the full range of state types.

@@

The transducer composition algorithm described above constructs a phylogenetic SCFG
for both ancestral and extant sequences.
A parse tree for this SCFG represents a structural and evolutionary
explanation of the extant sequences, including a complete ancestral reconstruction.
Consequently, given a set of extant sequences, many of the questions of interest to us
can be reduced to searches over, or summarizations of, the set of possible parse trees.

Well-known algorithms already exist for maxing or summing over SCFG parse tree likelihoods
(indeed, this is the motivation for our entire approach).
The Cocke-Younger-Kasami (CYK) algorithm performs maximum-likelihood (ML) inference;
the Inside algorithm can be used to sum over parse trees or sample them {\em a posteriori};
and the Inside-Outside algorithm yields posterior probabilities for individual parse tree nodes \cite{Durbin98}.

All of these algorithms are, however, complicated (at least in our models) by
the existence of ``null cycles'' in the grammar.
A null cycle is a parse tree fragment that is redundant and could be removed, such as a detour through $\Snull$ states ($A \to X \to Y \to X \to Y \to B$)
that could be replaced by a direct transition ($A \to B$).
Biologically, null cycles correspond to fragments of ancestral sequence that were universally deleted and therefore
are unobserved in any of the extant sequences.
These unobserved fragments can be unbounded in length (and so, therefore, can the parse tree).
Within the CYK, Inside and Outside recursions, this causes cyclic dependencies which cannot be resolved.

Below we describe a method to eliminate null cycles from
the ensemble model, by transforming any SCFG to an equivalent acyclic SCFG.
We then present multiple-sequence versions of the CYK, Inside and Outside algorithms.

As an aside: while some sort of null-cycle elimination is often required in order to deal with cyclic dependencies, there are several ways to accomplish this other than the algorithm presented here.
A simpler approach (that only works for the CYK algorithm) appears in the computational linguistics literature \cite{GecsegSteinby97}.
We have also developed a heuristic for CYK that simply ignores null cycles, as well as an iterative approximation that loops several times over cyclically-dependent cells of the DP matrix
until the estimate starts to converge. For conciseness, we have omitted descriptions of these methods, presenting only the exact elimination algorithm.

\subsubsection*{Exact elimination of null cycles in SCFGs}

@@

In contrast, an {\bf empty path} is defined as any parse-subtree {\em without} bifurcations that does not emit terminal symbols.
If $\Semit$ states $E_1$ and $E_2$ are connected in the state
graph via $\Snull$ states $N_1$ and $N_2$, then the path $E_1
\rightarrow N_1 \rightarrow N_2 \rightarrow E_2$ with  probability
$P(E_1 \rightarrow N_1) \cdot P(N_1 \rightarrow N_2) \cdot P(N_2
\rightarrow E_2)$ can be replaced by a single direct transition $E_1 \rightarrow
E_2$ with an identical probability.

@@

In our composite model, empty paths occur whenever a series of branch transducers ``winds back'' into $\Swait$ states.
Empty bifurcations occur when an entire substructure, present in an ancestor, is deleted in all that ancestor's extant descendants.

Empty paths and empty bifurcations are problematic because they can be combined to give finite-probability sequences of rules
that transform a nonterminal back into itself, with no observable emissions.
We refer to such sequences of rules as {\bf null cycles}.
As noted, null cycles generate cyclic dependencies in the CYK, Inside \& Outside algorithms.
Our goal is an algorithmic procedure to resolve these dependencies and account for the likelihood of such cycles by exact marginalization.

For simpler models, solutions to this problem are published.
Missing (empty) columns in independent-sites models can be accounted for by applying a correction factor $(1-p)^{-1}$
to account for the proportion of columns $p$ that are unobserved \cite{RivasEddy2008}.
The slightly more complicated situation of missing emissions in a HMM can be dealt with by
summing over all empty paths, yielding a geometric series that is solvable by matrix inversion
\cite{Holmes2003,Holmes2007,Lunter2007}.
Such algorithms
%, implemented by programs such as PhyloComposer \cite{Holmes2003,Holmes2007} and HMMoC \cite{Lunter2007},
effectively replace the HMM
with another HMM that contains no null cycles but is equivalent to the original, in that it models the
same probability distribution over sequences.
%The ``wing retraction'' algorithms used in the repetitively structured profile HMMs of HMMER \cite{hmmer} appear similar in operation.
However, these solutions do not easily generalize to SCFGs
(which may have empty bifurcations as well as empty paths).

@@

Note that step (i) is unique to SCFGs; step (ii), in contrast, is very similar to the empty-path elimination algorithm for HMMs.

We now give a formal description of the null-cycle elimination procedure,
along with procedures for probabilistically ``restoring'' null cycles
to sampled parse trees and Inside-Outside expectation counts.

@@

Exact inference on a star phylogeny with $N$ extant sequences therefore
has complexities $O(A \cdot L^{2N})$ and $O(B \cdot L^{3N})$ in memory and
time (respectively) for a multiple-SCFG with $A$ states and $B$ bifurcations. As
described earlier, in practice we frequently have expert knowledge
(such as a curated multiple alignment)
about the structures and/or evolutionary histories of the sequences of
interest.  We can use this knowledge as a constraint to reduce the accessible volume, and hence the storage requirements, of the DP matrix \cite{Holmes2005}.
Algorithms \ref{alg:inside}--\ref{alg:cyktraceback}, described below, give the Inside, CYK,
Outside and CYK traceback algorithms for a three-taxon star phylogeny,
constrained using the ``fold envelope'' concept.

@@

We illustrated our method for building models of structured sequences
using a simplified model which was introduced in previous work,
the TKF Structure Tree \cite{Holmes2004}, a simplified probabilistic
model of the evolution of RNA structure.

@@

The program that we wrote to simulate RNA structural alignments from the TKFST model, \evolsayer, is available as part of the same software distribution as \indiegram\ (the \dart\ package).
Another script (\animateevolsayer) can be used to make animations of the evolving RNA structures, using the RNAplot program in the \viennarna\ package \cite{HofackerEtAl94}.

@@

% Prior art
The work reported in this paper builds on extensive prior art in the areas of evolutionary modeling and ancestral reconstruction.
The reconstruction of ancient sequences was first proposed in 1963 by Pauling and Zuckerkandl \cite{PaulingZuckerkandl63};
current applications of this idea, mostly using substitution models, are surveyed in the book edited by Liberles \cite{Liberles2007}.
Many algorithms in phylogenetics implicitly reconstruct substitution histories,
whether by parsimony \cite{Edwards63,HendyPenny82} or likelihood \cite{Felsenstein81}.
There is a substantial body of work to model indels on phylogenies
\cite{ThorneEtal91,HeinEtal2000,Hein2001,HolmesBruno2001,SteelHein2001,Holmes2003,KnudsenMiyamoto2003,LunterSongMiklosHein2003,MiklosLunterHolmes2004,LunterEtAl2004,BradleyHolmes2007,SatijaEtAl2008}.
Recent work has extended these ideas to the reconstruction of indel
histories \cite{KimSinha2007,DialloEtAl2007},
particularly at the genomic scale \cite{MillerHausslerEtAl2006,PatenHolmesBirney2008}.
There is also prior work in computational linguistics
on the theory of transducers for sequences \cite{MohriPereiraRiley2000}
and parse trees \cite{Rounds70,Thatcher70,GecsegSteinby97,ComonEtAl2007-TreeTransducers}
(from which we take the terms ``string transducer'' and ``parse-tree transducer'').
We draw on the bioinformatics literature for SCFGs
\cite{Eddy94,Sakakibara94c,Durbin98},
especially Pair SCFGs \cite{RivasEddy99,Holmes2005,DowellEddy2006}
and phylogenetic SCFGs \cite{KnudsenHein2003}.
In particular, an early example of a pairwise conditional model $P(Y|X)$ for structure-dependent RNA evolution
may be found in \cite{KleinEddy2003}.
A conditional framework similar to ours in some respects is described
by Sakakibara et al \cite{Sakakibara2003}.
The dynamic programming inference algorithms for multiple-sequence SCFGs are closely related to the ``protosequence algorithm'' of Sankoff \cite{Sankoff85}.
%a seminal work in both RNA structural alignment and ancestral reconstruction.

@@

While we have focused on the TKF Structure Tree model in our Results,
our model-construction algorithm is applicable to any model of the evolution of secondary
structure which can be expressed as a Pair SCFG.
Realistic structural and thermodynamic effects---such as base-stacking
or loop length distributions---can, in principle, be incorporated.
%Furthermore, while an evolutionary model of structural evolution
%is necessarily complex,
%our model-construction algorithm enforces a compact parametrization,
%allowing fast training on structural alignments of pairs of sequences.
Other phenomena of RNA evolution may prove more difficult:
modeling helix slippage with a branch transducer is awkward,
let alone more radical changes in structure;
pseudoknots, too, are impossible with the models we have described here.
Even so, variants of our models could be used for proposing candidate alignments
for more accurate scoring by such models.

@@

Reconstructing structural changes of large RNAs
using the three-way sampling kernel which we have described
would require resources far in excess of those currently available;
barring the availability of supercomputers with terabytes of memory,
such algorithms will only be feasible for short RNAs (\tabref{complexities}).
A promising direction is to consider variations on the three-way sampling kernel,
such as the importance-sampling approach described for the TKF model
by \cite{RedelingsSuchard2005}.
This approach first proposes an ancestor $W$ by aligning extant
sequence $X$ to $Y$ (ignoring $Z$); then, in a second step, the proposed $W$ is
independently aligned to $Z$. The proposed three-way alignment and
reconstruction is then randomly accepted (or rejected) using a
Hastings ratio based on the three-way transducer composition. The
complexity of this kernel is the same as the pairwise case;
with suitable constraints, this is feasible for RNA
grammars on present hardware, at least for ribosomal domains (if not
yet whole subunits---although pairwise alignment of those should also
be possible soon). The approach of Redelings and Suchard therefore
merits future consideration in the context of modeling the evolution
of RNAs on a tree.

@@

While MCMC provides the most information about the posterior distribution of evolutionary histories,
in practice a maximum likelihood inference may be adequate (and typically faster).
The progressive profiling used by the \ortheus\ program for
reconstructing ancestral genomes is promising \cite{PatenHolmesBirney2008}.
This approach is similar to a progressive multiple alignment algorithm, in that it proceeds via a single postorder (leaf-to-root) traversal of the phylogeny.
As each node is visited, a profile is generated for that node, by aligning the profiles of its children to a composite transducer using DP, then sampling a finite number of traceback paths through the DP matrix.
The profile is not linear: the sampled paths instead form a reticulate network, a.k.a. a partial order graph \cite{LeeGrassoSharlow2002}.
An equivalent of \ortheus\ for RNA reconstruction should be possible,
representing the intermediate profiles using transducers.

In summary, the evolutionary models and algorithms we have described form a
systematic theoretical platform on which we can test different optimization and sampling strategies
for studying the structural evolution of RNA gene families in detail.
Stochastic grammars are powerful tools for this task; although they will not be the only tools we need,
particularly as we move towards modeling RNA evolution in greater detail,
they are flexible and serve our current purposes well.
Our hope is that these algorithms will allow us, and others, to test our
knowledge of RNA evolution by computational reconstruction and (eventually)
direct experimental investigation of early ribonucleic machines.
